---
title: "ST2195 Assignment 7"
author: "Lucas Rodriguez Cuitiño"
date: "2024-02-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load necessary libraries

```{r load-libraries, message=FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
```

## Read the dataset

```{r}
df <- read.csv("data/titanic.csv")

# Convert df columns to factors
df$Sex <- as.factor(df$Sex)
df$Pclass <- as.factor(df$Pclass)
df$Survived <- as.factor(df$Survived)
```

## Question 1: Bar charts to describe the gender, ticket class and survival of the passengers

### Bar chart for gender

```{r}
gender_plot <- ggplot(df, aes(x = Sex, fill = Sex)) +
  geom_bar() +
  labs(title = "Gender",
       x = "Gender",
       y = "Count") +
  theme_minimal()
gender_plot
```

### Bar chart for ticket class

```{r}
ticket_class_plot <- ggplot(df, aes(x = Pclass, fill = Pclass)) +
  geom_bar() +
  labs(title = "Ticket class",
       x = "Ticket class",
       y = "Count") +
  theme_minimal()
ticket_class_plot
```

### Bar chart for survival status

```{r}
survival_plot <- ggplot(df, aes(x = Survived, fill = Survived)) +
  geom_bar() +
  scale_x_discrete(labels = c("0" = "Not Survived", "1" = "Survived")) +
  labs(title = "Survival rate",
       x = "Survival status",
       y = "Count") +
  theme_minimal()
survival_plot
```

## Question 2: Passengers' age analysis by ticket class and survival status

### Histogram for passengers' age

```{r}
age_histogram <- ggplot(df, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  labs(title = "Age distribution",
       x = "Age",
       y = "Count") +
  theme_minimal()
age_histogram
```

```{r}
# Summarizing the Age column
age_summary <- summary(df$Age)

# Converting the summary to a list and creating a data frame
age_summary_df <- data.frame(as.list(age_summary))

# Printing the summary as a table
kable(age_summary_df, caption = "Summary of Age")
```

### Boxplot for age per Ticket Class

```{r}
age_ticket_class_boxplot <- ggplot(df, aes(x = Pclass, y = Age, fill = Pclass)) +
  geom_boxplot() +
  labs(title = "Age by ticket class",
       x = "Ticket class",
       y = "Age") +
  theme_minimal()
age_ticket_class_boxplot
```

```{r}
# Grouping by ticket class and summarizing age, including the number of NAs
age_ticket_class_summary <- df %>%
  group_by(Pclass) %>%
  summarize(
    Min = min(Age, na.rm = TRUE),
    `1st.Qu.` = quantile(Age, 0.25, na.rm = TRUE),
    Median = median(Age, na.rm = TRUE),
    Mean = mean(Age, na.rm = TRUE),
    `3rd.Qu.` = quantile(Age, 0.75, na.rm = TRUE),
    Max = max(Age, na.rm = TRUE),
    Non_NAs = sum(!is.na(Age)),
    NAs = sum(is.na(Age))
  )
kable(age_ticket_class_summary, caption = "Summary of Age by Ticket Class")
```

### Boxplot for age per Survival status

```{r}
age_ticket_class_boxplot <- ggplot(df, aes(x = Survived, y = Age, fill = Survived)) +
  geom_boxplot() +
  labs(title = "Age by survival status",
       x = "Survival status",
       y = "Age") +
  theme_minimal()
age_ticket_class_boxplot
```

```{r}
# Grouping by survival status and summarizing age
age_survival_summary <- df %>%
  group_by(Survived) %>%
  summarize(
    Min = min(Age, na.rm = TRUE),
    `1st.Qu.` = quantile(Age, 0.25, na.rm = TRUE),
    Median = median(Age, na.rm = TRUE),
    Mean = mean(Age, na.rm = TRUE),
    `3rd.Qu.` = quantile(Age, 0.75, na.rm = TRUE),
    Max = max(Age, na.rm = TRUE),
    Non_NAs = sum(!is.na(Age)),
    NAs = sum(is.na(Age))
  )

# Printing the summary as a table
kable(age_survival_summary, caption = "Summary of Age by Survival Status")
```

## Question 3: Travel fares and free passengers

### Histogram for Travel Fare

```{r}
fare_histogram <- ggplot(df, aes(x = Fare)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "black", na.rm = TRUE) +
  labs(title = "Travel fares",
       x = "Fare",
       y = "Count") +
  theme_minimal()
fare_histogram
```

```{r}
# Summarizing the Fare column
fare_summary <- summary(df$Fare)

# Converting the summary to a list and creating a data frame
fare_summary_df <- data.frame(as.list(fare_summary))

# Printing the summary as a table
kable(fare_summary_df, caption = "Summary of fares")
```

### Table with passenger fare payment status

```{r}
payment_summary <- df %>%
  mutate(Payment_Status = ifelse(Fare == 0, "Did Not Pay", "Did Pay")) %>%
  group_by(Payment_Status) %>%
  summarise(Count = n()) %>%
  ungroup()

knitr::kable(payment_summary, caption = "Passengers' Payment summary")
```

## Question 4: Different visualizations of family size per ticket class

```{r}
# FamilySize is the sum of siblings/spouses (SibSp), parents/children (Parch), plus one for the passenger themselves
df$FamilySize <- df$SibSp + df$Parch + 1
```

```{r}
# Summarizing each df$Column
sibsp_summary <- summary(df$SibSp)
parch_summary <- summary(df$Parch)
familysize_summary <- summary(df$FamilySize)

# Converting each summary to a list and creating a data frame for each
sibsp_summary_df <- data.frame(as.list(sibsp_summary))
parch_summary_df <- data.frame(as.list(parch_summary))
familysize_summary_df <- data.frame(as.list(familysize_summary))

# Binding the data frames together
combined_summary_df <- rbind(SibSp = sibsp_summary_df, Parch = parch_summary_df, FamilySize = familysize_summary_df)

# Printing the combined summary as a table
kable(combined_summary_df, caption = "Summary of SibSp, Parch, and FamilySize")
```

## Vertical boxplot

```{r}
family_size_boxplot_default <- ggplot(df, aes(x = Pclass, y = FamilySize, fill = Pclass)) +
  geom_boxplot() +
  labs(title = "Family size by ticket class",
       x = "Ticket class",
       y = "Family size") +
  theme_minimal()
family_size_boxplot_default
```

## Horizontal boxplot

```{r}
family_size_boxplot_horizontal <- ggplot(df, aes(y = Pclass, x = FamilySize, fill = Pclass)) +
  geom_boxplot() +
  labs(title = "Family size by ticket class",
       y = "Ticket class",
       x = "Family size") +
  theme_minimal()
family_size_boxplot_horizontal
```

## Vertical histogram

```{r}
family_size_histogram_default <- ggplot(df, aes(x = FamilySize, fill = Pclass)) +
  geom_histogram(binwidth = 1, color = "black") +
  facet_grid(~ Pclass) +
  labs(title = "Family size by ticket class",
       x = "Family size",
       y = "Count") +
  theme_minimal()
family_size_histogram_default
```

## Horizontal histogram

```{r}
family_size_histogram_horizontal <- ggplot(df, aes(x = FamilySize, fill = Pclass)) +
  geom_histogram(binwidth = 1, color = "black") +
  facet_wrap( ~ Pclass, ncol = 1) +
  labs(title = "Family size by ticket class",
       y = "Count",
       x = "Family size") +
  theme_minimal()
family_size_histogram_horizontal
```

## Vertical violin plot

```{r}
family_size_violin <- ggplot(df, aes(x = Pclass, y = FamilySize, fill = Pclass)) +
  geom_violin(trim = FALSE) + # Set trim=FALSE to show the full range of the data
  labs(title = "Family size by ticket class",
       x = "Ticket class",
       y = "Family size") +
  theme_minimal()
family_size_violin
```

## Horizontal violin plot

```{r}
family_size_violin_horizontal <- ggplot(df, aes(x = FamilySize, y = Pclass, fill = Pclass)) +
  geom_violin(trim = FALSE) + # Set trim=FALSE to show the full range of the data
  labs(title = "Family size by ticket class",
       x = "Family size",
       y = "Ticket class") +
  theme_minimal()
family_size_violin_horizontal
```
